---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# R-User-Group

<!-- badges: start -->

<!-- badges: end -->

This repository contains markdown files for topics we've discussed during the R user group. Feel free to peruse the index or search for posts using the GitHub search functionality.

Each topic has two files: a `.qmd` (Quarto) file and a `.md` (markdown) file. The `.md` file is rendered from the `.qmd` file. If you look through the `quarto-markdowns` folder you'll see one of each file for each topic.

------------------------------------------------------------------------

## Topics:

| Date     | Script                                 |
|----------| -------------------------------------  |
| 2025-05-29 | [APIs in R](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/apis-in-r.md)|
| 2025-05-15 | [saving data for downstream analyses](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/saving-data-for-downstream-analyses.md)
| 2025-04-03 | [confidence interval calculation](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/conf_int.md) |
| 2025-03-20 | [base R vs tidyverse](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/base-r_vs_tidyverse.md) |
| 2025-03-05 | [miscellaneous things](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/miscellaneous-things.md) |
| 2025-01-23 | [file management](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/file-management.md) |
| 2025-01-09 | [rendering multiple markdowns with parameters](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/paramaters_example.qmd) |
| 2024-10-31 | [intro to Quarto](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/intro-to-quarto.md) |
| 2024-10-17 | [`separate()` wider and longer by delim, regex, & position](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/separate-wider-and-longer.md) |
| 2024-10-03 | [`separate()` with delimiters](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/separate-by-delim.md) |
| 2024-03-22 | [file type comparison](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/file-type-comparison.md) |
| 2023-03-23 | [dynamically selecting variables](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/dynamic-selecting.md) |
| 2022-07-14 | [plotly basics](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/plotly-basics.md) |
| 2022-05-05 | [date basics](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/main/quarto-markdowns/date-basics.md) |

------------------------------------------------------------------------

## Contribution process:

If you'd like to add your R User Group presentation to this repo, please follow these steps:

1.  Create a .qmd document (HTML) and manually edit the YAML so it follows this outline:

```{r}
#| eval: false
---
format: gfm
title: "Title of document"
author: "Author1_First Author1_last, Author2_first Author2_last, etc."
date: "YYYY-MM-DD of the R User Group"
---
```

`format: gfm` tells R the output should be a "GitHub Flavored Markdown". A GFM is a basic HTML output, so some features (like including tab sets) aren't available but it generally meets our needs. Render your `. file so it generates an .md (markdown) file. This is the "finished" version.

2.  Once you've finished your Quarto doc and have pushed the `.qmd` and `.md` to GitHub, update the table on the `README.Rmd` file to include the date and a link to your new markdown. Documents should be added to the table in reverse chronological order (newest at the top). Render the `.Rmd` file and push the newly updated `.Rmd` and `.md` to GitHub.

Other things:

-   

    ## If you want to add images to your Quarto doc, save them in the `quarto-markdowns/images` folder. Prefix the image name with name of your .qmd file. Ex: the image for the `intro-to-quarto.qmd` doc is called `intro-to-quarto_visual-mode.png`.

## Future meetings:

2025-03-20: base R vs dplyr
