---
format: gfm
title: "packages for formatting tables"
author: "Eamonn Hartmann, Madelyn Sather, Sayema Badar, Beth Jump"
date: "2026-05-14"
execute: 
  warning: false
message: false
---
  
## Background

There are many packages you can use to format a table in Quarto. This document goes over 5 of them:

- [flextable](https://ardata-fr.github.io/flextable-book/)
- [GT](https://gt.rstudio.com/)
- [kable and kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
- [tinytable](https://vincentarelbundock.github.io/tinytable/) 
- [DT](https://rstudio.github.io/DT/)

In each section below, a different package is used to format the `palmerpenguins::penguins` data set. The code will go over how to: 

  - change the font type, font size and whether the font is regular, **bold** or *italic*
  - set column widths and heights
  - change the font color and the background of the cell
  - change the color and thickness of lines separating cells
  - format the header of the table
  - align text in a column (text should always be left aligned, numbers should always be right aligned)
  - merge cells vertically and horizontally. 

### Compatible formats

Generally, this is how when you should use these packages:

| package | save as PDF | HTML Quarto | PDF/Typst Quarto | Word Quarto |
| ------- | ----------- | ----------- | ---------------- | ----------- |
| `flextable` |     x    |             |       x          |     x       |
| `GT` | ... | ... | ... | ... |
| `kableExtra` | ... | ... | ... | ... |
| `tinytable` | ... | ... | ... | ... |
| `DT` | ... | ... | ... | ... |


```{r}

library(palmerpenguins)
library(tidyverse)

palmerpenguins::penguins

```

## `flextable`

```{r}

library(flextable)
library(officer)

```

### Clean data and create flextable

```{r}
penguins_table <- penguins %>%
  filter(!if_any(everything(), is.na)) %>%
  select(species, island, sex, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  mutate(sex = str_to_title(sex)) %>%
  head() %>%
  arrange(sex) %>%
  flextable()

```

### Change font type, size, bold, italics

For each of the following functions, use `i = ` to change the font parameters of a specific row and use `j = ` to change the font parameters of a specific column.

```{r}

penguins_table %>%  
  font(fontname = "Arial", part = "all") %>%
  fontsize(size = 12, part = "body") %>% 
  bold(bold = TRUE, part = "header") %>% 
  italic(italic = TRUE, part = "header") 

```
  
### Set cell width & height

For `width()`, use `j = ` to adjust the width of a specific column. For `height()`, use `i = ` to adjust the height of a specific row. 

```{r}
  
penguins_table %>%    
  width(width = 1.2, unit = "in") %>%
  height(height = 0.5, part = "header")

```

### Set font color 

The `officer` package is required to change the color of the font and background of cells and the thickness of the lines separating the cells.

```{r}
  
custom_line <- fp_border(color = "grey", style = "solid", width = 1)

penguins_table %>%  
  color(i = 6, j = 5, color = "red", part = "body") %>% 
  bg(i = 2, j = 4, bg = "yellow", part = "body") %>% 
  border_inner(border = custom_line)

```

### Set header labels

```{r}

penguins_table %>%
  set_header_labels(species = "Species", 
    island = "Island", bill_length_mm = "Bill length (mm)",
    bill_depth_mm = "Bill depth (mm)", flipper_length_mm = "Flipper length (mm)",
    body_mass_g = "Body mass (g)", sex = "Sex", year = "Year")

```

### Add row of labels to header

```{r}

penguins_table %>%
  add_header_row(values = c("Demographics", "Body measurements"), colwidths = c(3, 4)) %>%
  align(i = 1, part = "header", align = "left")

```

### Align contents within cell

```{r}

penguins_table %>%
  align(j = c(1:3), align = "left") %>%
  align(j = c(4:7), align = "right")

```
  
### Merge cells vertically

```{r}

penguins_table %>%
  merge_v(j = ~ species + island)

```

### Merge cells horizontally

The `merge_h()` function merges adjacent cells with duplicate values in the same row.

```{r}

penguins_table %>%
  merge_h()

```

### `autofit()`

The `autofit()` function can be used to adjust the size of cells to fit the size of the cell's contents.

```{r}

penguins_table %>%
  autofit()

```

## `GT`

```{r}
```

## `kable and kableExtra`

```{r}
```

## `tinytable`

```{r}
```

## `DT`

```{r}
```

