---
format: gfm
prefer-html: true
title: "packages for formatting tables"
author: "Eamonn Hartmann, Madelyn Sather, Sayema Badar, Beth Jump"
date: "2026-05-14"
execute: 
  warning: false
  message: false
---
  
## Background

There are many packages you can use to format a table in Quarto. This document goes over 5 of them:

- [flextable](https://ardata-fr.github.io/flextable-book/)
- [GT](https://gt.rstudio.com/)
- [kable and kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
- [tinytable](https://vincentarelbundock.github.io/tinytable/) 
- [DT](https://rstudio.github.io/DT/)

In each section below, a different package is used to format the `palmerpenguins::penguins` data set. The code will go over how to: 

  - change the font type, font size and whether the font is regular, **bold** or *italic*
  - set column widths and heights
  - change the font color and the background of the cell
  - change the color and thickness of lines separating cells
  - format the header of the table
  - align text in a column (text should always be left aligned, numbers should always be right aligned)
  - merge cells vertically and horizontally. 

### Compatible formats

Generally, this is how when you should use these packages:

| package | save as PDF | HTML Quarto | PDF/Typst Quarto | Word Quarto |
| ------- | ----------- | ----------- | ---------------- | ----------- |
| `flextable` |     x    |             |       x          |     x       |
| `GT` | ... | ... | ... | ... |
| `kableExtra` | ... | ... | ... | ... |
| `tinytable` | ... | ... | ... | ... |
| `DT` | ... | ... | ... | ... |


```{r}

library(palmerpenguins)
library(tidyverse)

penguins <- palmerpenguins::penguins

```

## `flextable`

```{r}

library(flextable)
library(officer)

```

### Clean data and create flextable

```{r}
penguins_table <- penguins %>%
  filter(!if_any(everything(), is.na)) %>%
  select(species, island, sex, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  mutate(sex = str_to_title(sex)) %>%
  head() %>%
  arrange(sex) %>%
  flextable()

```

### Change font type, size, bold, italics

For each of the following functions, use `i = ` to change the font parameters of a specific row and use `j = ` to change the font parameters of a specific column.

```{r}

penguins_table %>%  
  font(fontname = "Arial", part = "all") %>%
  fontsize(size = 12, part = "body") %>% 
  bold(bold = TRUE, part = "header") %>% 
  italic(italic = TRUE, part = "header") 

```
  
### Set cell width & height

For `width()`, use `j = ` to adjust the width of a specific column. For `height()`, use `i = ` to adjust the height of a specific row. 

```{r}
  
penguins_table %>%    
  width(width = 1.2, unit = "in") %>%
  height(height = 0.5, part = "header")

```

### Set font color 

The `officer` package is required to change the color of the font and background of cells and the thickness of the lines separating the cells.

```{r}
  
custom_line <- fp_border(color = "grey", style = "solid", width = 1)

penguins_table %>%  
  color(i = 6, j = 5, color = "red", part = "body") %>% 
  bg(i = 2, j = 4, bg = "yellow", part = "body") %>% 
  border_inner(border = custom_line)

```

### Set header labels

```{r}

penguins_table %>%
  set_header_labels(species = "Species", 
    island = "Island", bill_length_mm = "Bill length (mm)",
    bill_depth_mm = "Bill depth (mm)", flipper_length_mm = "Flipper length (mm)",
    body_mass_g = "Body mass (g)", sex = "Sex", year = "Year")

```

### Add row of labels to header

```{r}

penguins_table %>%
  add_header_row(values = c("Demographics", "Body measurements"), colwidths = c(3, 4)) %>%
  align(i = 1, part = "header", align = "left")

```

### Align contents within cell

```{r}

penguins_table %>%
  align(j = c(1:3), align = "left") %>%
  align(j = c(4:7), align = "right")

```
  
### Merge cells vertically

```{r}

penguins_table %>%
  merge_v(j = ~ species + island)

```

### Merge cells horizontally

The `merge_h()` function merges adjacent cells with duplicate values in the same row.

```{r}

penguins_table %>%
  merge_h()

```

### `autofit()`

The `autofit()` function can be used to adjust the size of cells to fit the size of the cell's contents.

```{r}

penguins_table %>%
  autofit()

```

## `GT`

```{r}

```

## `kable and kableExtra`

`kable` and `kableExtra` are a great option for making HTML tables. This [guide](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) includes tons of examples of customization for `kableExtra` tables!

### Clean data and create kable table
```{r}
library(kableExtra)

penguins_kable <- penguins %>%
  filter(!if_any(everything(), is.na)) %>%
  group_by(species, island, sex) %>%
  summarize(mean_bill_len = mean(bill_length_mm, na.rm = T),
            mean_flip_len = mean(flipper_length_mm, na.rm = T),
            mean_body_mass = mean(body_mass_g, na.rm = T),
            .groups = "keep") %>%
  ungroup() %>%
  kable()
```

### Change font type, size, bold, italics



```{r}
penguins_kable %>%
  kable_paper(html_font = "Cambria") %>%
  spec_font_size(x = 1)
```
  
### Set column widths & heights


```{r}

```

### Change font color and background of a cell


```{r}
```

### Change color and thickness of lines separating cells

```{r}
```

### Format the header of a the table

```{r}

```

### Align contents within cell

```{r}
penguins_kable 

```
  
### Merge cells vertically

```{r}
```

### Merge cells horizontally


```{r}

```
  

## `tinytable`

```{r}
```

## `DT`

The `DT` package creates interactive tables that the end user can search, filter and sort. 

```{r}
library(DT)

penguins_dt <- penguins %>%
  filter(species == "Adelie" &
           !is.na(sex)) %>%
  group_by(species, island, sex) %>%
  summarize(mean_bill_len = mean(bill_length_mm, na.rm = T),
            mean_flip_len = mean(flipper_length_mm, na.rm = T),
            mean_body_mass = mean(body_mass_g, na.rm = T),
            .groups = "keep") %>%
  ungroup() %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))
```

### Change font type, size, bold, italics

You can change font size, weight and italics using the `formatStyle()` function. You need to use JavaScript notation to change the font (see `list(initComplete())`). 

You can try to do any of these things with JavaScript notation. See [here](https://stackoverflow.com/questions/44101055/changing-font-size-in-r-datatables-dt) for how to change font size using Javascript.

```{r}
penguins_dt %>%
  datatable(options = list(
    initComplete = JS(
      "function(settings, json) {",
      "$('body').css({'font-family': 'Calibri'});",
      "}"))) %>%
  formatStyle(columns = 'island', fontSize = '20%') %>%
  formatStyle(columns = 'sex', fontWeight = 'bold') %>%
  formatStyle(columns = 'species', fontStyle = 'italic')
```

### Set column widths & heights

*Note: the internet says this should work but it's not. Come back!!*

```{r}
penguins_dt %>%
  datatable(options = list(
    autoWidth = FALSE,
    columnDefs = list(list(targets = "_all", width = '20px'))))
```

### Change font color and background of a cell

```{r}
penguins_dt %>%
  datatable() %>%
  formatStyle('species',  
              color = '#2C92B8') %>%
  formatStyle('sex',
              backgroundColor = '#FFDF20')

```

### Change color and thickness of lines separating cells

I couldn't find an easy way to edit the color and thickness of lines separating cells. Instead I've included information on how to generally style a table. 

Here is a [guide](https://datatables.net/manual/styling/classes) to the options available for styling a DT table. Just list all of the options you want to enable in the `class` argument. Options should be separated by a space, not a comma. In the example below our table has cell borders, is compact and striped. 

```{r}
penguins_dt %>%
  datatable(class = 'cell-border compact stripe')
```

### Format the header of the table

Here is how to change the text color and background and make the font bold using JavaScript notation. 

```{r}
penguins_dt %>%
  datatable(options = list(
  initComplete = JS(
    "function(settings, json) {",
    "$(this.api().table().header()).css({'background-color': '#000', 'color': '#ffce00', 'font-style': 'bold'});",
    "}")
))


```

Here is how to make a totally custom header (taken from [here](https://rstudio.github.io/DT/)).

```{r}
penguins_dt <- penguins_dt %>%
  select(island, everything())

sketch <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'Island'),
      th(colspan = 2, 'Species & sex'),
      th(colspan = 3, 'Mean measurements')
    ),
    tr(
      lapply(colnames(penguins_dt[2:ncol(penguins_dt)]), th)
    )
  )
))

penguins_dt %>%
  select(island, everything()) %>%
  datatable(container = sketch, rownames = F)

```

### Align contents within cell

Here we're centering all (`_all`) headers, left aligning text columns and right aligning numerical columns. 

```{r}
penguins_dt %>%
  datatable(options = list(columnDefs = list(
    list(targets = "_all", className = "dt-head-center"), # Center header text
    list(targets = 1:3, className = 'dt-left'),
    list(targets = 4:6, className = 'dt-right'))))
```
  
### Merge cells horizontally

Here's how to use values from a column to group output by value. Note: the grouping column is specified by the number here, `list(dataSrc = c(1))`. Unlike normal R things, the indexing here starts at 0, not 1. So if you wanted to group by the first column you'd use a 0, the second column you'd use a 1, etc. 

I haven't figured out how to change the column alignment of the grouped variable. 

```{r}
penguins_dt %>%
  datatable(rownames = FALSE, 
           extensions = 'RowGroup', 
           options = list(rowGroup = list(dataSrc = c(1)),
                          columnDefs = list(list(targets = 1, visible = FALSE))))
         
```

### Merge cells horizontally

This [stack overflow](https://stackoverflow.com/questions/39484118/shiny-merge-cells-in-dtdatatable) post goes over how to do this in Shiny but I don't fully get it. 

### Bonus: exporting `datatable` data

If you include a `DT` table in a Quarto doc or R Shiny dashboard, you can allow end users to download the data right from your doc! [Source](https://www.r-bloggers.com/2021/05/datatable-editor-dt-package-in-r/).

```{r}
penguins_dt %>%
  datatable(extensions = 'Buttons',
            options = list(dom='Bfrtip',
                           buttons=c('copy', 'csv', 'excel', 'print', 'pdf')))
```