---
format: gfm
prefer-html: true
title: "packages for formatting tables"
author: "Eamonn Hartmann, Madelyn Sather, Sayema Badar, Beth Jump"
date: "2026-05-14"
execute: 
  warning: false
  message: false
---
  
## Background

There are many packages you can use to format a table in Quarto. This document goes over 5 of them:

- [flextable](https://ardata-fr.github.io/flextable-book/)
- [GT](https://gt.rstudio.com/)
- [kable and kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
- [tinytable](https://vincentarelbundock.github.io/tinytable/) 
- [DT](https://rstudio.github.io/DT/)

In each section below, a different package is used to format the `palmerpenguins::penguins` data set. The code will go over how to: 

  - change the font type, font size and whether the font is regular, **bold** or *italic*
  - set column widths and heights
  - change the font color and the background of the cell
  - change the color and thickness of lines separating cells
  - format the header of the table
  - align text in a column (text should always be left aligned, numbers should always be right aligned)
  - merge cells vertically and horizontally. 

### Compatible formats

Generally, this is how when you should use these packages:

| package | save as PDF | HTML Quarto | PDF/Typst Quarto | Word Quarto |
| ------- | ----------- | ----------- | ---------------- | ----------- |
| `flextable` |     x    |             |       x          |     x       |
| `GT` |     x    |     x      |       x          |             |
| `kableExtra` | ... | ... | ... | ... |
| `tinytable` | ... | ... | ... | ... |
| `DT` | ... | ... | ... | ... |


```{r}

library(palmerpenguins)
library(tidyverse)

penguins <- palmerpenguins::penguins

```

## `flextable`

```{r}

library(flextable)
library(officer)

```

### Clean data and create flextable

```{r}
penguins_table <- penguins %>%
  filter(!if_any(everything(), is.na)) %>%
  select(species, island, sex, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  mutate(sex = str_to_title(sex)) %>%
  head() %>%
  arrange(sex) %>%
  flextable()

```

### Change font type, size, bold, italics

For each of the following functions, use `i = ` to change the font parameters of a specific row and use `j = ` to change the font parameters of a specific column.

```{r}

penguins_table %>%  
  font(fontname = "Arial", part = "all") %>%
  fontsize(size = 12, part = "body") %>% 
  bold(bold = TRUE, part = "header") %>% 
  italic(italic = TRUE, part = "header") 

```
  
### Set cell width & height

For `width()`, use `j = ` to adjust the width of a specific column. For `height()`, use `i = ` to adjust the height of a specific row. 

```{r}
  
penguins_table %>%    
  width(width = 1.2, unit = "in") %>%
  height(height = 0.5, part = "header")

```

### Set font color 

The `officer` package is required to change the color of the font and background of cells and the thickness of the lines separating the cells.

```{r}
  
custom_line <- fp_border(color = "grey", style = "solid", width = 1)

penguins_table %>%  
  color(i = 6, j = 5, color = "red", part = "body") %>% 
  bg(i = 2, j = 4, bg = "yellow", part = "body") %>% 
  border_inner(border = custom_line)

```

### Set header labels

```{r}

penguins_table %>%
  set_header_labels(species = "Species", 
    island = "Island", bill_length_mm = "Bill length (mm)",
    bill_depth_mm = "Bill depth (mm)", flipper_length_mm = "Flipper length (mm)",
    body_mass_g = "Body mass (g)", sex = "Sex", year = "Year")

```

### Add row of labels to header

```{r}

penguins_table %>%
  add_header_row(values = c("Demographics", "Body measurements"), colwidths = c(3, 4)) %>%
  align(i = 1, part = "header", align = "left")

```

### Align contents within cell

```{r}

penguins_table %>%
  align(j = c(1:3), align = "left") %>%
  align(j = c(4:7), align = "right")

```
  
### Merge cells vertically

```{r}

penguins_table %>%
  merge_v(j = ~ species + island)

```

### Merge cells horizontally

The `merge_h()` function merges adjacent cells with duplicate values in the same row.

```{r}

penguins_table %>%
  merge_h()

```

### `autofit()`

The `autofit()` function can be used to adjust the size of cells to fit the size of the cell's contents.

```{r}

penguins_table %>%
  autofit()

```

## `GT`

The **gt** package is a great package to use for tables when your format is HTML. It offers a wide variety of styling options and is really great for when you want to dynamically style the table based on the table contents. There are many different ways to style tables with the **gt** package, so please see (https://gt.rstudio.com/) for a reference guide on different styling options.

```{r}

library(gt)

```

### Clean data and create GT table

```{r}

penguins_table_gt <- penguins %>%
  filter(!if_any(everything(), is.na)) %>%
  select(species, island, sex, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  mutate(sex = str_to_title(sex)) %>%
  head() %>%
  arrange(sex) %>%
  gt::gt()

```

### Change font type, size, bold, italics

The `opt_table_font()` function can be used to change font type, size, bold, and italics for the entire table. 

```{r}

penguins_table_gt %>%
  opt_table_font(font = "Arial",
                 size = "12px",
                 weight = "bold",
                 style = "italic")

```

The `tab_style()` and `locations()` functions can be used to change font type, size, bold, and italics for specific rows/columns, the table body, or the table header.

```{r}


penguins_table_gt %>%
  
  # table headers
  tab_style(style = cell_text(size = "16px", 
                              weight = "bold"),
            locations = cells_column_labels(columns = everything())) %>%
  
  # table body
  tab_style(style = cell_text(size = "9px"),
            locations = cells_body()) %>%
  
  # specific row
  tab_style(style = cell_text(style = "italic"),
            locations = cells_body(rows = 1)) %>%
  
  # specific column
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_body(columns = bill_length_mm))


```

### Set cell width & height

To set widths and heights for the entire table, use `tab_options()` function. To set widths for specific columns, use the `cols_width()` function.

```{r}

penguins_table_gt %>%
  
  # Set specific column widths
  cols_width(bill_length_mm ~ px(125)) %>%
  
  # Set row height (affects all rows)
  tab_options(data_row.padding = px(0),
              table.width = px(1000))

```

### Set font color 

The `tab_style()` and `locations()` functions can be used to change font or background color for specific rows/columns and cells.

```{r}

penguins_table_gt %>%
    tab_style(style = cell_fill(color = "orange"),
              locations = cells_body(rows = bill_length_mm > 40)) %>%
    
  tab_style(style = cell_text(color = "red"),
            locations = cells_body(columns = bill_length_mm)) %>%
  
  tab_style(style = cell_fill(color = "yellow"),
            locations = cells_body(columns = bill_depth_mm,
                                  rows = bill_depth_mm > 20))

```

### Set header labels

Use the `cols_label` to change the column names.

```{r}

penguins_table_gt %>%
  cols_label(species = "Species", 
             island = "Island", 
             bill_length_mm = "Bill length (mm)",
             bill_depth_mm = "Bill depth (mm)", 
             flipper_length_mm = "Flipper length (mm)",
             body_mass_g = "Body mass (g)", 
             sex = "Sex")

```

### Add row of labels to header and titles

Use the `tab_header()` function to add table titles. Use the `tab_spanner()` function to create grouped column headers (spanners). 

```{r}

penguins_table_gt %>%
  tab_header(title = md("**Penguins Table**"),
             subtitle = "GT Table") %>%
  tab_spanner(label = "Demographics",
              columns = c("species", "island", "sex")) %>%
  tab_spanner(label = "Body Measurements",
              columns = c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"))

```

### Align contents within cell

Use the `cols_align()` function to align the contents within a cell.

```{r}

penguins_table_gt %>%
  cols_align(align = "center",
             columns = everything()) %>%
  cols_align(align = "left",
             columns = c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"))


```

## `kable and kableExtra`

`kable` and `kableExtra` are a great option for making HTML tables. This [guide](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) includes tons of examples of customization for `kableExtra` tables!

### Clean data and create kable table

```{r}
library(kableExtra)

penguins_kable <- penguins %>%
  filter(!if_any(everything(), is.na)) %>%
  group_by(island, species, sex) %>%
  summarize(mean_bill_len = mean(bill_length_mm, na.rm = T),
            mean_flip_len = mean(flipper_length_mm, na.rm = T),
            mean_body_mass = mean(body_mass_g, na.rm = T),
            .groups = "keep") %>%
  ungroup() %>%
  kable()
```

### Change font type, size, bold, italics

You can set the `font_size` in the kable theme or in `row_spec()` but it doesn't work if you set it in `column_spec()`

```{r}
penguins_kable %>%
  kable_paper(html_font = "Cambria", font_size = 16) %>%
  row_spec(1:3, bold = TRUE, font_size = 10) %>%
  column_spec(4:6, italic = TRUE)
```
  
### Set column widths & heights

You can easily set column widths with the `width` argument. Making the columns taller is a bit less straightforward but using the `extra_css = "padding: 10px"` works. You can change the "10px" to however many pixels or inches you want. 

```{r}
penguins_kable %>%
  kable_classic() %>%
  column_spec(1:2, width = c('30%', '30%')) %>%
  row_spec(1:4,
           extra_css = "padding: 10px")
```

### Change font color and background of a cell

`row_spec()` and `column_spec()` allow you to color specific rows and columns by position. 

```{r}
penguins_kable %>%
  kable_minimal() %>%
  row_spec(1:3, color = "#2C92B8") %>%
  column_spec(4:6, background = "darkgrey", color = "white")
```

### Change color and thickness of lines separating cells

The easiest way to do this is by using the different themes (`kable_minimal`, `kable_paper`, etc) that are built into the `kableExtra` package. If you want to change the lines on a specific cell, you can try playing around with LaTeX as proposed [here](https://stackoverflow.com/questions/77647118/how-to-add-horizontal-lines-in-kable-and-the-bottom-line-being-thicker-and-orang). 

### Format the header of a the table

In `kableExtra` you can format the header with the `row_spec()` function by referring to the top row as 0. 

```{r}
penguins_kable %>%
  row_spec(0, font_size = 18, bold = T, italic = T)
```

Here's how to add an additional header above the table. 

```{r}
penguins_kable %>%
  add_header_above(c("", "Species & sex" = 2, "Avg measurements" = 3), bold = TRUE)
```

### Align contents within cell

By default, `kableExtra` aligns text to the left and numbers to the right, which is excellent!! If you want to align all of the cells in the table, you can use `align = "c"` (or `"l"` or `"r"`) in the `kbl()` function. 

```{r}
penguins[1:4, 1:4] %>%
  kable(align = 'rccl')
```
  
### Merge cells vertically

The `collapse_rows()` function makes this super easy!

```{r}
penguins_kable %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, valign = "top")
```

### Merge cells horizontally

It doesn't seem like you can merge columns in `kableExtra` but you can group sets of rows with `pack_rows()` which might come in handy. 

```{r}
penguins_kable %>%
  pack_rows(index = c("Group 1" = 4,
                      "Group 2" = 4, 
                      "Group 3" = 2),
            label_row_css = "background-color: #666; color: #fff;")
```
  
### Bonus: adding plots into the table

`kableExtra` makes it pretty easy to add tiny plots to your table! See [here](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html#Insert_Images_into_Columns) (you need to scroll down a bit to the section just before `Row Spec`) for an example! 

## `tinytable`

```{r}
```

## `DT`

The `DT` package creates interactive tables that the end user can search, filter and sort. These don't render in a GitHub-flavored markdown so I've just displayed the code below. If you want to test these out, you can render this on your local machine and you'll be able to see the tables. 

```{r}
#| eval: false
library(DT)

penguins_dt <- penguins %>%
  filter(species == "Adelie" &
           !is.na(sex)) %>%
  group_by(species, island, sex) %>%
  summarize(mean_bill_len = mean(bill_length_mm, na.rm = T),
            mean_flip_len = mean(flipper_length_mm, na.rm = T),
            mean_body_mass = mean(body_mass_g, na.rm = T),
            .groups = "keep") %>%
  ungroup() %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))
```

### Change font type, size, bold, italics

You can change font size, weight and italics using the `formatStyle()` function. You need to use JavaScript notation to change the font (see `list(initComplete())`). 

You can try to do any of these things with JavaScript notation. See [here](https://stackoverflow.com/questions/44101055/changing-font-size-in-r-datatables-dt) for how to change font size using Javascript.

```{r}
#| eval: false

penguins_dt %>%
  datatable(options = list(
    initComplete = JS(
      "function(settings, json) {",
      "$('body').css({'font-family': 'Calibri'});",
      "}"))) %>%
  formatStyle(columns = 'island', fontSize = '20%') %>%
  formatStyle(columns = 'sex', fontWeight = 'bold') %>%
  formatStyle(columns = 'species', fontStyle = 'italic') %>%
  kbl(escape = FALSE) %>%
  kable_paper() %>%
  save_kable(file = "table1.html", self_contained = F)

```

### Set column widths & heights

```{r}
#| eval: false

penguins_dt %>%
  datatable(options = list(
    autoWidth = FALSE,
    columnDefs = list(list(targets = "_all", width = '20px'))))
```

### Change font color and background of a cell

```{r}
#| eval: false

penguins_dt %>%
  datatable() %>%
  formatStyle('species',  
              color = '#2C92B8') %>%
  formatStyle('sex',
              backgroundColor = '#FFDF20')

```

### Change color and thickness of lines separating cells

I couldn't find an easy way to edit the color and thickness of lines separating cells. Instead I've included information on how to generally style a table. 

Here is a [guide](https://datatables.net/manual/styling/classes) to the options available for styling a DT table. Just list all of the options you want to enable in the `class` argument. Options should be separated by a space, not a comma. In the example below our table has cell borders, is compact and striped. 

```{r}
#| eval: false

penguins_dt %>%
  datatable(class = 'cell-border compact stripe')
```

### Format the header of the table

Here is how to change the text color and background and make the font bold using JavaScript notation. 

```{r}
#| eval: false

penguins_dt %>%
  datatable(options = list(
  initComplete = JS(
    "function(settings, json) {",
    "$(this.api().table().header()).css({'background-color': '#000', 'color': '#ffce00', 'font-style': 'bold'});",
    "}")
))


```

Here is how to make a totally custom header (taken from [here](https://rstudio.github.io/DT/)).

```{r}
#| eval: false

penguins_dt <- penguins_dt %>%
  select(island, everything())

sketch <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'Island'),
      th(colspan = 2, 'Species & sex'),
      th(colspan = 3, 'Mean measurements')
    ),
    tr(
      lapply(colnames(penguins_dt[2:ncol(penguins_dt)]), th)
    )
  )
))

penguins_dt %>%
  select(island, everything()) %>%
  datatable(container = sketch, rownames = F)

```

### Align contents within cell

Here we're centering all (`_all`) headers, left aligning text columns and right aligning numerical columns. 

```{r}
#| eval: false

penguins_dt %>%
  datatable(options = list(columnDefs = list(
    list(targets = "_all", className = "dt-head-center"), # Center header text
    list(targets = 1:3, className = 'dt-left'),
    list(targets = 4:6, className = 'dt-right'))))
```
  
### Merge cells horizontally

Here's how to use values from a column to group output by value. Note: the grouping column is specified by the number here, `list(dataSrc = c(1))`. Unlike normal R things, the indexing here starts at 0, not 1. So if you wanted to group by the first column you'd use a 0, the second column you'd use a 1, etc. 

I haven't figured out how to change the column alignment of the grouped variable. 

```{r}
#| eval: false

penguins_dt %>%
  datatable(rownames = FALSE, 
           extensions = 'RowGroup', 
           options = list(rowGroup = list(dataSrc = c(1)),
                          columnDefs = list(list(targets = 1, visible = FALSE))))
         
```

### Merge cells horizontally

This [stack overflow](https://stackoverflow.com/questions/39484118/shiny-merge-cells-in-dtdatatable) post goes over how to do this in Shiny but I don't fully get it. 

### Bonus: exporting `datatable` data

If you include a `DT` table in a Quarto doc or R Shiny dashboard, you can allow end users to download the data right from your doc! [Source](https://www.r-bloggers.com/2021/05/datatable-editor-dt-package-in-r/).

```{r}
#| eval: false

penguins_dt %>%
  datatable(extensions = 'Buttons',
            options = list(dom='Bfrtip',
                           buttons=c('copy', 'csv', 'excel', 'print', 'pdf')))
```