---
title: "Tidy evaluation"
format: gfm
author: "Beth Jump"
date: "2025-09-04"
execute:
  warning: false
---

## Background

For some background on tidy evaluation (a type of non-standard evaluation) check out Jenny Bryant's talk from 2019 and this overview of [Programming with `dplyr`](https://dplyr.tidyverse.org/articles/programming.html#introduction). Essentially, tidy evaluation is what lets us reference variables without reference to their data frame in `tidyverse` functions. 

I'm not going to get into the technical details here - I'll just go over how it intersects with our work and how to deal with it. 

Consider the `palmerpenguins::penguins` data set:

```{r}
library(tidyverse)
data <- palmerpenguins::penguins
glimpse(data)
```

`data` is an object in our environment that contains these data. We can reference `data` without quotes because it's an object.

The `species` column is a part of the `data` object but it does not exist as its own object in our R environment. To reference the column, we need to tell R how to find it. We can do this the base R way where we explicilty reference the data:

```{r}
data$species
```

Or we can use tidy evaluation with the `tidyverse` functions:

```{r}
data %>%
  select(species)
```

## In practice

Generally speaking, we don't need to know *how* tidy evaluation works, we just need to know that it works with `tidyverse` functions. The place where it breaks down is when we want to write functions that use `tidyverse` syntax and want to pass variables as arguments. 


