---
title: "`tidyr::unite()`"
format: gfm
author: "Beth Jump"
date: "2025-09-04"
execute:
  warning: false
---

## Background

We've talked about `tidyr::separate()` in the past, `tidyr::unite()` is the inverse of `tidyr::separate()`. I typically use `tidyr::unite()` when I want to paste the contents of multiple columns into a single variable. 

## Examples

### Pasting contents of specific columns

Here is a data frame of parts of emails. We have columns for `username`, `domain` and `extension`:

```{r}
library(tidyverse)

emails <- data.frame(
  username = c("hogwarts14", "hpotter", "wildraccoon101", "rweasley"),
  domain = c("yahoo", "hogwarts", "gmail", "hogwarts"),
  extension = c(".com", ".edu", ".com", ".edu")
)
emails
```

We can use `tidyr::unite()` to neatly combine these into a single variable `email`. 

```{r}
emails %>%
  mutate(at = "@") %>% 
  unite(col = "email",
        c("username", "at", "domain", "extension"),
        sep = "")
```

You could also use `paste()` or `paste0()` within a `mutate()` step to combine these, it really comes down to preference. One pro of using `unite()` is that it will automatically drop the variables used in the `unite()` step. 

### Pasting contents of a set of columns

This is where `unite()` really proves its worth. 

Let's say someone requests a file that has a row for each penguin species and island and single column that lists the weights of the penguins in that category from small to high. 