---
format: gfm
title: "Base R statistics-related functions"
author: "Beth Jump"
date: "2025-06-26"
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse)
```
  
## Background

R was originally created for statisticians and there are some very handy functions in base R and in the `stats` package (which is included with the R installation).

## `sample()`

The `sample()` function allows you to select a sample from a list of numbers. This is helpful both for randomly picking numbers from an array and for generating a fake set of observatons from a distribution. Consider these two examples:

* We want to randomly select three unique numbers between 0-100: 

```{r}
sample(x = 0:100,
       size = 3)
```
Note that we aren't using the `replace` argument because by default `replace = FALSE`. If we didn't care about selecting the same number twice, we could set `replace = TRUE`. 

* We want to generate some fake data from a set of numbers


```{r}
dist <- sample(x = 0:5,
               size = 1000,
               replace = T,
               prob = c(10, 9, 5, rep(1, 3)))
```


## `rep()`

The `rep()` function just lets you repeat and element a given number of times. Let's say you want to make a variable and you need the value `"yellow"` included three times. Instead of typing: `c("yellow", "yellow", "yellow")` you can use:

```{r}
rep("yellow", 3)
```

You might not use this very often, but if you find yourself typing the same thing over and over consider using `rep()` instead!

## `replicate()`

`rep()` just repeats the same element or result over and over while `replicate()` will actually re-run code a specified number of times. We used `replicate()` when we [compared the size, read and save times of different types of files](https://github.com/San-Mateo-County-Health-Epidemiology/R-User-Group/blob/a40a9f5eec3c8f3b3ba4b83cb5d0cfb35124ab8e/quarto-markdowns/r-scripts/file-type-comparison.R). 

Here's a quick copmar




## `_norm()`

```{r}

```


## `_binom()`

```{r}

```
