---
format: gfm
title: "Base R statistics-related functions"
author: "Beth Jump"
date: "2025-06-26"
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse)
```
  
## Background

R was originally created for statisticians and there are some very handy functions in base R and in the `stats` package which is included with the R installation.

## `rep()`

```{r}

```


## `sample()`

The `sample()` function allows you to select a sample from a list of numbers. This is helpful both for randomly picking numbers from an array and for generating a fake set of observatons from a distribution. Consider these two examples:

* We want to randomly select three unique numbers between 0-100: 

```{r}
sample(x = 0:100,
       size = 3)
```
Note that we aren't using the `replace` argument because by default `replace = FALSE`. If we didn't care about selecting the same number twice, we could set `replace = TRUE`. 

* We want to generate some fake data from a set of numbers


```{r}
dist <- sample(x = 0:5,
               size = 1000,
               replace = T,
               prob = c(10, 9, 5, rep(1, 3)))
```



## `_norm()`

```{r}

```


## `_binom()`

```{r}

```
