---
title: "Comparing code running time"
author: "Beth Jump"
date: "2022-10-06"
format: gfm
execute: 
  warning: false
message: false
---

## Background

When you're processing a lot of data at once it's important to try to optimize the speed of your code. If there is a section of your code that can be written in different ways, you should compare the methods and, when possible, use the option that runs the fastest.

## Comparing single runs 

Here we have some fake data that contains 10,000 rows of two dates:

```{r}
library(rbenchmark)
library(tidyverse)

data <- data.frame(
  date1 = seq.Date(Sys.Date() - days(9999), Sys.Date(), 1),
  date2 = rep(Sys.Date(), 1000)
)

head(data, 10)
```

We want to calculate the number of days between these two dates in the most efficient way possible. We know three different ways to do this:

1. `base::difftime()`
2. `lubridate::interval()`
3. simple math: date2 - date1

We want to know which method is fastest. We can do this by timing each method and comparing how long each one takes:

### `base::difftime()`

```{r}
start <- Sys.time()
data1 <- data %>%
  mutate(date_diff = difftime(date2, date1, units = "days"))
end <- Sys.time()

difftime_time <- end-start
difftime_time
```

### `lubridate::interval())`

```{r}
start <- Sys.time()
data1 <- data %>%
  mutate(date_diff = interval(date2, date1) %/% days(1))
end <- Sys.time()

interval_time <- end-start
interval_time
```
### simple math

```{r}
start <- Sys.time()
data1 <- data %>%
  mutate(date_diff = date2 - date1)
end <- Sys.time()

math_time <- end-start
math_time
```
Looking at all of the times together, we see that `lubridate::interval()` is the slowest, followed by `base::difftime()` and basic math:

```{r}
#| echo: false
data.frame(
  difftime_time, interval_time, math_time
)
```

But can we be confident in that? Probably not. We need to compare more than one run to figure out which method is quickest. 

## Comparing multiple runs

It's always better to compare hundreds or thousands of runs than just compare one. We can use the `rbenchmark::benchmark()` function to compare multiple runs at once!

*Note: there are other packages that also do this, I am most familiar with this one.*

Here we are comparing each method across 1,000 runs. 

```{r}
#| cache: true
benchmark("difftime" = {
  data1 <- data %>%
    mutate(date_diff = difftime(date2, date1, units = "days"))
},
"lubridate" = {
  data1 <- data %>%
    mutate(date_diff = interval(date2, date1) %/% days(1))
},
"math" = {
  data1 <- data %>%
    mutate(date_diff = date2 - date1)
},
replications = 1000,
columns = c("test", "replications", "elapsed",
            "relative", "user.self", "sys.self"))
```

When we compare 1,000 runs, we confirm that `lubridate::interval()` is the slowest, but we also see that `base::difftime()` is slightly faster than simple math. 