---
title: "Comparing code running time"
author: "Beth Jump"
date: "2022-10-06"
format: gfm
execute: 
  warning: false
message: false
---

## Background

When you're processing a lot of data at once, it's important to try and optimize the speed of your code. If there is a secion of your code that can be written in different ways, you should compare the methods and, when possible, use the one that runs the fasted. 
The first step to doing this is knowing how long your code takes to run. 

## Comparing single runs 

Here we have some fake data that is just two dates:

```{r}
library(rbenchmark)
library(tidyverse)

data <- data.frame(
  date1 = seq.Date(Sys.Date() - days(9999), Sys.Date(), 1),
  date2 = rep(Sys.Date(), 1000)
)

head(data, 10)
```

We want to calculate the time between these two dates in the most efficient way possible. We know three different ways to do this:

1. `base::difftime()`
2. `lubridate::interval()`
3. simple math: date2 - date1

We want to know which is fastest! We can do this the long way by using each method and comparing how long each one takes:

### `base::difftime()`

```{r}
start <- Sys.time()
data1 <- data %>%
  mutate(date_diff = difftime(date2, date1, units = "days"))
end <- Sys.time()

difftime_time <- end-start
difftime_time
```

### `lubridate::interval())`

```{r}
start <- Sys.time()
data1 <- data %>%
  mutate(date_diff = interval(date2, date1) %/% days(1))
end <- Sys.time()

interval_time <- end-start
interval_time
```
### simple math

```{r}
start <- Sys.time()
data1 <- data %>%
  mutate(date_diff = date2 - date1)
end <- Sys.time()

math_time <- end-start
math_time
```

```{r}
data.frame(
  difftime_time, interval_time, math_time
)
```

## Comparing multiple runs

By comparing one run, we can see that the `lubridate::interval()` method is ~2 times slower than the `base::difftime()` and simple math methods. But, we don't know if those results will hold true every time. It's better to compare hundreds or thousands of runs than to just compare one. 

We can use the `rbenchmark::benchmark()` function to compare multiple runs!

```{r}
#| cache: true
benchmark("difftime" = {
  data1 <- data %>%
    mutate(date_diff = difftime(date2, date1, units = "days"))
},
"lubridate" = {
  data1 <- data %>%
    mutate(date_diff = interval(date2, date1) %/% days(1))
},
"math" = {
  data1 <- data %>%
    mutate(date_diff = date2 - date1)
},
replications = 1000,
columns = c("test", "replications", "elapsed",
            "relative", "user.self", "sys.self"))
```

When we compare 1,000 runs, we confirm that `lubridate::interval()` is the slowest, but we also see that `base::difftime()` is slightly faster than simple math. 