---
format: gfm
title: "`dplyr` practice part 1"
author: "Beth Jump"
date: "2022-11-17"
execute: 
  warning: false
message: false
---

## Background

[`dplyr`](https://dplyr.tidyverse.org/) is a package within the [`tidyverse`](https://tidyverse.org/) ecosystem that is focused on basic data wrangling.

## Examples

We're going to walk through how to solve some basic data tasks using functions from the `dplyr` package using the [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/) data set. 

```{r}
library(dplyr)
data <- palmerpenguins::penguins_raw
```

There are six questions and example solutions below. If your solution code is different but you get to the same answer, that is totally fine!

### 1. How many unique studies are in this data set and how many penguins are in each?

You can use the `count()` or `distinct()` function to get the names of unique studies, but only `count` will tell you how many records are in each study. `distinct()` will only give you the study names.

```{r}
data %>%
  count(studyName)
```

A more verbose way to solve this would be with `group_by()` and `summarize()`:

```{r}
data %>%
  group_by(studyName) %>%
  summarize(penguin_ct = n()) %>%
  ungroup()
```

### 2. Which penguin (use `Individual ID`) weighs the least (`Body Mass (g)`) in each study?

Here can use a combination of `group_by()`, `arrange()` and `slice()`. 

```{r}
data %>%
  group_by(studyName) %>%
  arrange(`Body Mass (g)`) %>%
  slice(1) %>%
  ungroup() %>%
  select(studyName, `Individual ID`, `Body Mass (g)`)
```

### 3. What are the average and median flipper lengths of the all penguins in each study?

Like the previous question, you need to use `group_by()` but here you also should use the `summarize()` function. 

```{r}
data %>%
  filter(!is.na(`Flipper Length (mm)`)) %>%
  group_by(studyName) %>%
  summarize(penguin_ct = n(),
            mean_flipper = mean(`Flipper Length (mm)`),
            median_flipper = median(`Flipper Length (mm)`)) %>%
  ungroup()


```

### 4. Create a table where within each species, penguins are numbered from largest Culmen Length to smallest.

If you use the `row_number()` function within a `mutate()` step, it will number records. Make sure you've sorted your data with `arrange()` so that your numbers are assigned as expected.  

```{r}
data %>%
  group_by(Species) %>%
  arrange(desc(`Culmen Length (mm)`)) %>%
  mutate(record_num = row_number()) %>%
  ungroup() %>%
  select(Species, `Culmen Length (mm)`, record_num) %>%
  arrange(record_num, Species)
```

### Bonus 1: Which island had the most penguins and which has the earliest egg?

`n()` and `min()` within a summarize function will give you the number per group and the minimum of a given value respectively. 

```{r}
data %>%
  group_by(Island) %>%
  summarize(penguin_ct = n(),
            earliest_egg = min(`Date Egg`, na.rm = T)) %>%
  ungroup()
```

### Bonus 2: Create a table that shows the breakdown by sex (count and percent) by island.

You don't need to format the `percent` variable, but you can using the `percent()` function from the [`scales`](https://scales.r-lib.org/) package. 

```{r}
data %>%
  count(Island, Sex) %>%
  group_by(Island) %>%
  mutate(total = sum(n),
         pct = scales::percent(n/total, accuracy = 0.1))
```
