---
format: gfm
title: "Functions - part 2"
author: "Beth Jump"
date: "2022-09-22"
execute:
  eval: FALSE
---

## Overview

This is a continuation of `functions - part 1` from 2022-05-19 and goes over additional options when writing custom functions. 

This is the basic syntax for writing a function. Here we have one input and the function just returns that input:

```{r}
new_function <- function(x){
  x
}

new_function("hello")
```

You can add as many inputs as you need:

```{r}
new_function1 <- function(x, y, z){
  x*y/z
}

new_function1(6, 5, 2)
new_function1(z = 2, y = 5, x = 6)
```

## Additional options

### Specifying default arguments

To make the function a bit easier for users, you can specify defaults for arguments. Here we have defaults for each argument so the user can run the function without specifying anything:

```{r}
new_function2 <- function(x = 5, y = 6, z = 2){
  x*y/z
}
new_function2()
```

The user can easily override defaults by specifying their own value for an argument:
`# you can also override the defaults!
```{r}
new_function2(z = 1)
```

### Passing variables as arguments 

#### Technical things

If you want to specify a variable as an argument in a function that uses `tidyverse` or `ggplot2` syntax, you need to use some special functions and syntax. These include:

   - `!!`: "bang bang" operator. A good explanation of how to use that is [here](https://www.r-bloggers.com/2019/07/bang-bang-how-to-program-with-dplyr/)
   - `quo()`: explanation of quosures [here](https://www.geeksforgeeks.org/r-language/quosures-in-r-language/)
   - `!!sym()`: explanation of `sym()` [here](https://www.rdocumentation.org/packages/rlang/versions/1.1.6/topics/sym)
   - `get()`: returns the value of a names object. More [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/get)
   
Whenever I'm working with variables as arguments, I play around to see which of these options works best in the context of my function. 

#### Example

Let's say we want to make a function to make a chart with `ggplot2`. First, we'll write the code for the chart:
```{r}
library(tidyverse)

data <- palmerpenguins::penguins

ggplot(data, 
       aes(x = body_mass_g, 
           y = flipper_length_mm)) +
  geom_point()
```

If we put this into a function where the variables are referenced as objects, the function won't work:
```{r}
#| eval: false
gg_dot <- function(data, x, y){
  ggplot(data, 
         aes(x = x, y = y)) +
    geom_point()
}

gg_dot(data, body_mass_g, flipper_length_mm)
```

To make this work, we need to make two changes: 

1. We will precede the variables that can be specified in the function with `!!`
2. When we run the function, we will wrapped the specified variables with `quo()`


```{r}
gg_dot <- function(data, x, y){
  ggplot(data, aes(x = !!x, y = !!y)) +
    geom_point()
}

gg_dot(data, quo(body_mass_g), quo(flipper_length_mm))
```

### Validating inputs within the functions

If your function is going to be widely used, you might want to consider 
