---
format: gfm
title: "Use here package to call and save files"
author: "Julie Bartels"
date: "2026-02-04"
execute:
  eval: TRUE
---

## Overview

This tutorial explains how to use the *here* package to call, run, and save files. This is useful because as long as you are within an R project, you don't need to use long file paths, which can be inconvenient, provide information about how your files are stored, and are not necessarily replicable across users. 

## Here package

You will need to load the **here** package:
```{r}
# install.packages('here')
library(here)
```
When you run this, you see that the package tells you within which directory (folder), you are operating. In our case, the last part of the directory file path should read "/R-User-Group".

### Using the here package to write/save/load dataframes

As long as the directory you want to save to is located within your overall project directory ("/R-User-Group"), you can write your dataframe using the **here** package.

The structure of the code is as follows.

- Within the *write_csv()* function, call *here()* within the **here** package. Note: I am using here::here, because there is another package with a *here()* function, so I want to ensure I am calling the *here()* within the **here** package.

- Starting with the directory below the main project folder, include the name of the directory in quotation marks, followed by a comma. For the following example file path ("/R-User-Group/data/raw/hospitalizations), you would write ("data", "raw", "hospitalizations").

- Then once you have listed all the necessary folders, you include the name of the specific file (.csv, .xlsx, .RDS) within quotation marks.
 
For example, we will create a csv of the palmer penguins dataset and save it to data (and then delete it).

### Side note: pacman package

This is a fun tip - when you need to load a bunch of packages, you can use the **pacman** package. This is just a little bit of a cleaner look to your code.

```{r, output = FALSE}
# Load pacman package
# install.packages('pacman')
library(pacman)

# Load packages to prepare the data and conduct analysis
# install.packages('palmerpenguins')
p_load(dplyr, tidyverse, palmerpenguins)
```

```{r}
# Make dataframe of penguin data
penguin_data <- palmerpenguins::penguins

head(penguin_data)
```

```{r}
# Save penguin dataframe to data folder in /R-User-Group
penguin_data %>%
  write_csv(here::here("data", "penguin_data.csv"))
```

If you check the data folder, it should be there - we can then use the **here** package to load this csv, just for fun:

```{r, output = FALSE}
# Remove penguin_data from our global environment
rm(penguin_data)

# Load penguin_data from the data folder
penguin_data <- read_csv(here::here("data", "penguin_data.csv"))
```
Now we will remove the penguin dataframe from the folder, also using the *here* package. 

```{r, output = FALSE}
file.remove(here::here("data", "penguin_data.csv"))
```
Finally, we clean up the global environment - this is just good practice to keep things tidy!

```{r}
rm(penguin_data)
```


