---
title: "looking at data"
format: gfm
author: "Beth Jump"
date: "2022-06-16"
execute:
  warning: false
---

## Overview

This goes through some basic methods for looking at data. These methods should be used any time you get a new data set. 

```{r}
library(tidyverse)

data <- palmerpenguins::penguins
```

## Methods

We'll focus on some basic methods for reviewing and cleaning your data. We'll look at:

- generating summary statistics for a variable
- looking at unique values for a variable
- de-duplicating data

### Summary statistics 

The `dplyr::summarize()` function is great for looking at the distribution of variables in your data set. You can add a `group_by()` statement before the `summarize()` function to get summary statistics for specific values within a variable.

Here is one way to get the distribution of `body_mass_g` by penguin species:

```{r}
data %>%
  group_by(species) %>%
  summarize(total = n(),
            min = min(body_mass_g, na.rm = T),
            median = median(body_mass_g, na.rm = T),
            mean = mean(body_mass_g, na.rm = T),
            sd = sd(body_mass_g, na.rm = T),
            max = max(body_mass_g, na.rm = T),
            nas = sum(is.na(body_mass_g))) %>%
  ungroup()
```

Base R also has a useful function called `summary()` that will give you the numeric distribution of a variable with a lot less code than `summarize()`. Here's the distribution of `body_mass_g` for all penguins:

```{r}
summary(data$body_mass_g)
```
`summary()` also works with categorical variables:

```{r}
summary(data$species)
```

### Unique values

You can use `dplyr::distinct()` and `dplyr::count()` to look at the values within a variable. `distinct()` returns only the values while `count()` returns the values and their frequencies. If you're working with a large data set, `count()` will take longer to run than `distinct()`. 

#### distinct()

```{r}
data %>%
  distinct(species)
```

#### count()

```{r}
data %>%
  count(species)
```

### De-duplicating data 

#### Identifying duplicates

You should always check for duplicates in your data. There are many ways to do this, but my favorite is with `group_by()` and `filter(n() > 1)`. `filter(n() > 1)` will keep any groups that have more than one observation in them. If this filter returns an empty data set, then you know that the selection of variables in your `group_by()` statement will identify unique observations. 

Can we identify unique penguins by only looking at `species`, `island`, `year`, `sex` and `body_mass_g`? 

```{r}
data %>%
  group_by(species, island, year, sex, body_mass_g) %>%
  filter(n() > 1) %>%
  arrange(species, island, year, sex, body_mass_g)
```

No. All of the rows in the table are duplicates. 


Can we identify unique penguins by looking at `species`, `island`, `year`, `sex`, `body_mass_g`, `flipper_length_mm` and `bill_length_mm`? 

```{r}
data %>%
  group_by(species, island, year, sex, body_mass_g, flipper_length_mm, bill_length_mm) %>%
  filter(n() > 1) %>%
  arrange(species, island, year, sex, body_mass_g)
```

We can!!

#### De-duplicating methods

You can de-duplicate data indiscriminately with `distinct()` or you can do it a bit more carefully with a combination of `group_by()`, `arrange()` and `slice()` or if you need to combine groups, you can use `summarize()`. 

If we only want to keep one penguin of each species on each island, but don't care which penguin we keep, we can use `distinct()`. 

```{r}
data %>%
  distinct(species, island, .keep_all = T)
```

If we're doing downstream analyses, we probably want to select our penguins a bit more carefully. We can do this with `group_by()`, `arrange()` and `slice()`. Here's how we would keep the penguin with the smallest bill length for each species and island:

```{r}
data %>%
  group_by(species, island) %>%
  arrange(bill_length_mm) %>%
  slice(1) %>%
  ungroup()
```

If we want to calculate summary statistics about sub-groups, we can use `summarize()` to get these values and de-duplicate our data. Here's how to get the average bill length for each species and island:

```{r}
data %>%
  group_by(species, island) %>%
  summarize(avg_bill_length = mean(bill_length_mm, na.rm = T),
            .groups = "keep") %>%
  ungroup()

```