---
format: gfm
title: "Multi-row Axis Labels in ggplot"
author: "Madelyn Sather"
date: "2026-02-19"
---

## Overview

The **legendry** package provides additional guide functionality to the **ggplot2** ecosystem. More information about the **legendry** package can be found [here](https://www.rdocumentation.org/packages/legendry/versions/0.2.1) and [here](https://teunbrand.github.io/legendry/). In this document, we will go over how to add multi-row axis labels to your plots using the **legendry** package. This is especially helpful for date variables.

### Load Libraries

```{r warning=F, message=F}
library(ggplot2)
library(dplyr)
library(lubridate)
library(smcepi)
library(legendry) #devtools::install_github("teunbrand/legendry")
```

### Load Data

```{r}
data <- ggplot2::economics
glimpse(data)
```

### Clean Data

Let's add some date variables that we might want to show on our axis.

```{r}
data_clean <- data %>%
  mutate(year = year(date),
         qtr = quarter(date))
```

## Create a Plot with Multi-row Axis Labels

### Interaction Function

To create multi-row axis labels, you'll need to include `interaction()` inside of `aes()` for whichever axis you want your multi-row label to be on. `interaction()` computes a factor which represents the interaction of the given factors and creates a combined factor for plotting. The result is always unordered.

```{r message=F}
graph_by_qtr <- data_clean %>%
  filter(year %in% 1977:1987) %>%
  group_by(qtr, year) %>%
  summarize(tot_unemploy = sum(unemploy)) %>%
  ungroup() %>%
  ggplot(aes(x = interaction(qtr, year),
             y = tot_unemploy)) + 
  geom_col(fill = "steelblue") +
  labs(title = "**Unemployment by quarter (1977-1987)**",
       x = "Quarter",
       y = "Number of Unemployed") +
  scale_y_continuous(limits = c(0, NA),
                     expand = c(0, 0)) +
  theme_gg_smc()

```

### Guide Axis Nested Function

Next, you'll also want to include `guide_axis_nested()` from the **legendry** package inside of `guides()`. This function is what creates the multi-level (nested) axis labels. You'll need to specifiy which axis you want your nested axis on. More information about `guide_axis_nested()` can be found [here](https://www.rdocumentation.org/packages/legendry/versions/0.2.1/topics/guide_axis_nested). 

Below is an example graph of the number of unemployed by quarter over a period of ten years. Here I've included `type = "fence"`, which draws vertical lines between groups, and `rail = "inner"`, which draws a horizontal line between groups.

```{r}
graph_by_qtr + guides(x = guide_axis_nested(type = "fence", 
                                            rail = "inner"))
```

### Theme Guide Function

You can also play around with different themes using `theme_guide()` from the **legendry** package. `theme_guide()` allows you to customize the appearance of the axes using theme elements. More information on the theme_guide() function can be found [here](https://www.rdocumentation.org/packages/legendry/versions/0.2.1/topics/theme_guide).

Here are some of example graphs of different ways to style the nested axis using `theme_guide()`.

```{r}
graph_by_qtr + guides(x = guide_axis_nested(type = "box")) +
  theme_guide(box = element_rect("limegreen", "forestgreen"))
```

```{r}
graph_by_qtr + guides(x = guide_axis_nested(type = "fence", rail = "inner")) +
  theme_guide(
    fence.post = element_line("tomato"),
    fence.rail = element_line("dodgerblue")
  )
```
