---
format: gfm
title: "parameters_example"
author: Hanley Kingston
date: "2025-01-09"
params:
  person: !expr c("Simba", "Mowgli", "Mufasa")
  message: "Hello"
  happy_tone: TRUE
---

When run as a stand-along script, the parameters specified in the yaml will be loaded as named elements in a list named params.

## Save params as their own objects

```{r}
person <- params$person
message <- params$message
happy_tone <- params$happy_tone
```

## Write string using parameter inputs

```{r}
cat(paste0(person, ",\n", message, ifelse(happy_tone == TRUE, "!", "."), "\n\n"))
```

## Rendering using a separate script

This markdown can also be rendered through a separate script using:

```{r, eval = FALSE}
rmarkdown::render(
  "parameters_example.qmd",
  output_file = "<output_file>.md",
  params = list(
    person = "<person(s)>",
    message = "<message(s)>",
    happy_tone = "<T/F>"),
  envir = new.env())
```

For any parameter not specified, the value in the .qmd yaml will be used as a default.

For an example of rending this markdown with different inputs passed through a for-loop, see: r-scripts//render_parameters_example.R

## Other notes

-   Do not use rm(list = ls()). This will erase the parameters passed in the yaml

-   For R Markdown, use !r instead of !expr to pass an object to params
