---
title: "strings and regular expressions"
format: gfm
author: "Beth Jump"
date: "2023-03-28"
execute:
  warning: false
---

## Overview

If you're working with string variables, I recommend leaning heavily on the `stringr` package and using regular expressions. The [`stringr` cheat sheet](https://rstudio.github.io/cheatsheets/strings.pdf), is a great overview of the package and also has some regular expression shortcuts. 

Regular expressions are a method for identifying patterns in strings. We'll review some below and you can also practice regular expressions with this [tutorial](https://regexone.com/).

`separate()` and `unite()` from the `tidyr` package are also helpful for splitting or combining string variables in data sets. 

When dealing with string variables, I recommend testing out the functions on a list or variable before you try using it within a mutate step. It's easier to play and test things out on a couple of variables instead of a giant vector that is part of a large data set.

## String manipulation

We'll use these data to run through some string cleaning tasks below. 

```{r}
library(tidyverse)

data <- data.frame(
  phones = c("650-555-9023", "(415)123-6789", "6509984567", "(650)4456374"),
  emails = c("hogwarts14@yahoo.com", "hpotter@hogwarts.edu", "wildraccoon101@gmail.com", "rweasley@hogwarts.edu")
)
data
```

#### 1. Format the phone numbers so that they all follow the same pattern. 

```{r}
data1 <- data %>%
  mutate(phones = str_replace_all(phones, "\\-|\\(|\\)", ""))
data1
```

#### 2. Copy the phone area code into a new variable called `area_code`. 

```{r}
data2 <- data1 %>%
  mutate(area_code = str_sub(phones, 1, 3))
data2
```

#### 3. Create a new variable called `hogwarts_student` and flag the `hogwarts.edu` emails.

```{r}
data3 <- data2 %>%
  mutate(hogwarts_student = case_when(str_detect(emails, "hogwarts.edu$") ~ 1, 
                                      TRUE ~ 0))
data3
```

#### 4. Split the email into three variables, `user`, `domain` and `extension`. 

```{r}
data4 <- data3 %>%
  separate(emails, 
           c("user", "domain", "extension"),
           sep = "\\@|\\.")
data4
```