---
format: gfm
title: "when and how to use factors"
author: "Beth Jump"
date: "2026-04-16"
execute: 
  warning: false
message: false
---

## Background

Factors are a way to store your string/character data as an ordered list. An interesting (and brief) history of factors in R is [here](https://simplystats.github.io/2015/07/24/stringsasfactors-an-unauthorized-biography/). 

The [`forcats`](https://forcats.tidyverse.org/) package is a tidyverse package for dealing with factors. 

When would you want to store your data in factors?

* If you need to sort by a categorical value in a way that is not alphabetical (ex: by month)
* If you're using a string variable in a mathematical model
* If there is a limited number of categories
* ??

When would it be better to store your character data as a string?

* If you need to manipulate the contents of the string
* If your character variable isn't easily grouped (ex: a notes field, a list of first names, etc.)
* If you're assigning factors based on other variables in an automated or semi-automated script where data might change. You can still use a factor, but you should assign the values explicitly
* ??

```{r}
#| echo: false

library(forcats)
library(lubridate)
library(dplyr)
```

## Motivating example

Let's say we have a list of months. We want to sort these chronologically. If we sort the character variable, we end up with an *alphabetical* list:

```{r}
month_list <- c("March", "July", "April", "August")
sort(month_list)
```

Instead, we can make this into a factor and assign the levels according to the chronological order of months. Then our data will sort correctly!

```{r}
levels <- as.character(month(1:12, abbr = F, label = T))

month_list_factor <- factor(x = month_list,
                            levels = levels)

sort(month_list_factor)
```

## Useful functions

### making a factor

Using the `base::factor()` function is the most straightforward way to make a factor but it can also be a pain if your factor has many levels. Luckily, the `forcats` package has some functions that can make creating factors quicker and more dynamic!

#### by a different variable

If you want one variable to be ordered by a different variable, you can use `fct_reorder()`. 

Here we are sorting colors by the length of each color (ex: blue = 4, yellow = 6)

```{r}
data.frame(
  colors = c("blue", "orange", "yellow", "green", "red")
) %>%
  mutate(letter_count = nchar(colors),
         color_fct = fct_reorder(colors, letter_count)) %>%
  arrange(color_fct)
```

Note that "yellow" and "orange" have the same number of letters so they are ordered somewhat arbitrarily. You could override this by assigning levels manually. 

More examples are [here](https://forcats.tidyverse.org/reference/fct_reorder.html). 

#### by a frequency 

`forcats` has a very handy function called `fct_infreq()` that will order your data from least to most frequent. 

```{r}
data <- data.frame(
  colors = sample(c("blue", "orange", "yellow", "green", "red"), 1000, replace = T, prob = (5:1))
) %>%
  group_by(colors) %>%
  mutate(count = n()) %>%
  ungroup()

data %>%
  mutate(colors = fct_infreq(colors)) %>%
  arrange(colors) %>%
  distinct(colors, count)

```

As far as I can tell there is no `fct_freq()` (to order things from large to small), but the internet suggests just wrapping `fct_infreq()` in `fct_rev()` which reverses the factors: 

```{r}
data %>%
  mutate(colors_fct = fct_rev(fct_infreq(colors))) %>%
  arrange(colors_fct) %>%
  distinct(colors, colors_fct, count)
```

#### by a frequency for a chart

In data visualizations, sometimes we can't show all categories. There is a really fun function called `fct_lump` that will lump smaller groups together!

```{r}
data1 <- data %>%
  mutate(colors_fct = fct_lump(colors, n = 2))
   
levels(data1$colors_fct)

```
For more info and functions, check out the [cheat sheet](https://rstudio.github.io/cheatsheets/html/factors.html).